---
title: "AHS_2011"
output: html_document
---

```{r load_AHS}
library(glptools)
glp_load_packages(T)

AHS_2011_micro <- read.csv("C:/Users/morga/OneDrive - The Community Foundation of Louisville, Inc/accessible_housing/data/ahs2011m.csv")
```
Harvard metrics = NOSTEP, 
NOSTEP = Entrance accessible without steps or stairs
BUILT = year built
      Response Codes
      1919: 1919 or earlier
      1920: 1920 to 1929
      1930: 1930 to 1939
      1940: 1940 to 1949
      1950: 1950 to 1959
      1960: 1960 to 1969
      1970: 1970 to 1974
      1975: 1975 to 1979
      1980: 1980 to 1984
      1985: 1985 to 1989
      1990: 1990 to 1994
      1995: 1995 to 1999
      2000 : YYYY: 2000 to YYYY
NUNITS = Number of living units in building (1-997)
NUNIT2 = Structure type
CANE = Someone in household uses a cane or walker
CRUTCH = Someone in household uses crutches
ECHAIR = Someone in the household uses an electric wheelchair, cart, or scooter
WCHAIR = Someone in the household uses a manual wheelchair
HDSB = Household has a disabled person
STAIRS = stairs in unit
STAIRRL = Stair railings firmly attached
HMACAB = Presence and use of wheelchair accessible kitchen cabinets
HMBRL = Presence and use of handrails or grab bars in bathroom
HMBROOM = Presence and use of wheelchair accessible bathroom
HMBST = Presence and use of built-in shower seats
HMCLCTRL = Presence and use of wheelchair accessible climate controls
HMCOUNT = Presence and use of wheelchair accessible countertops
HMELEVATE = Presence and use of elevators in home
HMENTBD = Presence and use of bedroom on entry level
HMENTBTH = full bath entry
HMHNDLE = handles instead of knobs
HMHNDRL = handrails or grab bars
HMKIT - wheelchair acc kitchen
HMLEVEL = no steps between rooms
HMOUTLET = wheelchair accessible outlets
HMORL =  handrails or grab bars in other areas
HMRAMPS = ramps in home
HMSKLVR= sink handles or levers
HMSWITCH = Presence and use of wheelchair accessible electrical switches
HMWHEELN = Number of wheelchair accessible features in the home
HMXWDR = Presence and use of extra-wide doors or hallways

```{r}
accessible_features_2011 <- AHS_2011_micro %>%
  select(NOSTEP, CANE, CRUTCH, ECHAIR, WCHAIR, HDSB, STAIRS, STAIRRL, HMACAB, HMBRL, HMBROOM, HMBST, HMCLCTRL, HMCOUNT, HMELEVATE, HMENTBD, HMENTBTH, HMHNDRL, HMKIT, HMLEVEL, HMOUTLET, HMORL, HMRAMPS, HMSKLVR, HMSWITCH, HMWHEELN, HMXWDR, BUILT, NUNITS, NUNIT2, WEIGHT)%>%
  mutate(across(everything(), ~ as.integer(str_replace_all(str_trim(.), "'", "")))) %>%
  mutate(
    NOSTEP = case_when(
      NOSTEP %in% c(-9, -6) ~ NA, 
      NOSTEP == 1 ~ T,
      NOSTEP == 2 ~ F),
    
    STAIRRL = case_when(
      STAIRRL %in% c(-9, -6) ~ NA, 
      STAIRRL == 1 ~ TRUE, # only true when all stairs have attached rails
      STAIRRL %in% c(2,3,4) ~ FALSE)
    ) %>%
  mutate(across(c(WCHAIR, ECHAIR, CANE, CRUTCH, HDSB, STAIRS),
                ~ case_when(.x %in% c(-6, -9) ~ NA,
                            .x == 1 ~ TRUE, 
                            .x == 2 ~ FALSE,
                            TRUE ~ NA)
                )) %>%
  mutate(across(c(HMACAB, HMBRL, HMBROOM, HMBST, HMCLCTRL, HMCOUNT, HMELEVATE, HMENTBD, HMENTBTH, HMHNDRL, HMKIT, HMLEVEL, HMOUTLET, HMORL, HMRAMPS, HMSKLVR, HMSWITCH, HMWHEELN, HMXWDR), 
                ~ case_when(.x %in%  c(-6, -9) ~ NA,
                            .x %in% c(1,2) ~ TRUE, # 1/2 = is present, is/is not in use 
                            .x == 3 ~ FALSE, 
                            TRUE ~ NA)
                )) %>%
  mutate(
    SF = case_when(
      NUNIT2 %in% c(1,2, 4) ~ TRUE, 
      NUNIT2 %in% c(3) ~ FALSE), 
    MF = case_when(
      NUNIT2 == 3 ~ TRUE,      
      NUNIT2 %in% c(1,2,4) ~ FALSE),
    
    yr_built = case_when(
      BUILT <= 1940 ~ "built_1939_or_earlier" , 
      BUILT %in% 1940:1950 ~ "built_1940_to_1959", 
      BUILT %in% 1960:1975 ~ "built_1960_to_1979", 
      BUILT %in% 1980:1995 ~ "built_1980_to_1999", 
      BUILT >= 2000 ~ "built after 2000"
    ))


```