---
title: "projections"
output: html_document
---
###2019 AHS Variables Measuring Accessibility: 
NOSTEP - Flag indicating unit's entrance is accessible without climbing up or down any stairs
UNITFLOORS - Number of floors within the unit
HABATHENTRY- Flag indicating bathroom on the entry level (when there's more than 1 floor)
HABEDENTRY - Flag indicating bedroom on the entry level (when there's more than 1 floor)
HALIFT- Flag indicating home has a chairlift, stairlift, or platform lift
HARAMP- Flag indicating home has a ramp

```{r load_AHS}

library(glptools)
glp_load_packages(T)

AHS_2019_micro <- read.csv("C:/Users/morga/OneDrive - The Community Foundation of Louisville, Inc/data/ahs2019n.csv")
```

```{r clean_filter_AHS}

# 1: Yes
# 2: No
# M or -9: Not reported
# N or -6: Not applicable

accessible_features <- AHS_2019_micro %>%
  select(NOSTEP, UNITFLOORS, HABATHENTRY, HABEDENTRY, HALIFT, HARAMP, WEIGHT) %>%
  mutate(across(everything(), ~ as.integer(str_replace_all(str_trim(.), "'", "")))) %>%
  mutate(
    no_step = case_when(
      NOSTEP == 9 ~ NA, 
      NOSTEP == 1 ~ T,
      NOSTEP == 2 ~ F),
    
    entry_bath = case_when(
      HABATHENTRY == 9 ~ NA, 
      HABATHENTRY == 1 ~ T,
      HABATHENTRY == 2 ~ F, 
      HABATHENTRY == 6 ~ T # single floor unit 
      ), 
    
    entry_bed = case_when(
      HABEDENTRY == 9 ~ NA, 
      HABEDENTRY == 1 ~ T,
      HABEDENTRY == 2 ~ F, 
      HABEDENTRY == 6 ~ T # single floor unit 
      ), 
    
    lift = case_when(
      HALIFT == 9 ~ NA, 
      HALIFT == 1 ~ T,
      HALIFT == 2 ~ F, 
      HALIFT == 6 ~ T # single floor unit or not applicable  
      ), 
    
    ramp = case_when(
      HARAMP == 9 ~ NA, 
      HARAMP == 1 ~ T,
      HARAMP == 2 ~ F, 
      HARAMP == 6 ~ T # single floor unit or not applicable  
      ), 
    
    floors = case_when(
      UNITFLOORS == 1 ~ "1",
      UNITFLOORS == 2 ~ "2",
      UNITFLOORS == 3 ~ "3 or more"
    )
    )

save(accessible_features, file = "interim_data/accessible_features_hh.RData")
```

```{r accessibility_summary}
accessibility_summary <- accessible_features %>%
  #all individual features
  summarize(
    no_step_pct = weighted.mean(no_step, WEIGHT, na.rm = TRUE) * 100,
    entry_bath_pct = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE) * 100,
    entry_bed_pct = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE) * 100,
    lift_pct = weighted.mean(lift, WEIGHT, na.rm = TRUE) * 100,
    ramp_pct = weighted.mean(ramp, WEIGHT, na.rm = TRUE) * 100, 
    
    #basic accessibility
    core_accessibility_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1,
    WEIGHT,
    na.rm = TRUE
  ) * 100,

  # basic features plus ramp OR lift
  core_plus_ramp_lift_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1 & (ramp == 1 | lift == 1),
    WEIGHT,
    na.rm = TRUE
  ) * 100
  )

accessibility_summary <- accessibility_summary %>%
  pivot_longer(everything(), names_to = "feature", values_to = "percent")
  
save(accessibility_summary, file = "interim_data/accessibility_summary.RData")
```

```{r ACS_housing_estimates}
library(readxl)
library(dplyr)
library(tidyr)

jeff_co_HU <- read_excel("data/CO-EST2024-HU-21.xlsx")

jeff_co_HU <- jeff_co_HU %>%
  janitor::row_to_names(4, remove_rows_above = TRUE) %>%
  janitor::clean_names() %>%
  rename(estimates_base = april_1_2020_estimates_base) %>%
  mutate(estimates_base = as.integer(estimates_base)) %>%
  filter(geographic_area == ".Jefferson County, Kentucky") %>%
  pivot_longer(estimates_base:x2024, names_to = "year", values_to = "total_units")

```

```{r project!}
jeff_acc_housing_est <- jeff_co_HU %>%
  filter(year == "x2024") %>% #most recent estimates
  crossing(accessibility_summary) %>%     # Cartesian join: every year × every metric
  mutate(
    #round down to nearest whole num
    estimated_units = floor(total_units * (percent / 100))
  ) 

save(jeff_acc_housing_est, file = "interim_data/accessible_housing_estimates.RData")

```
