---
title: "projections"
output: html_document
---
###2019 AHS Variables Measuring Accessibility: 
NOSTEP - Flag indicating unit's entrance is accessible without climbing up or down any stairs
UNITFLOORS - Number of floors within the unit
HABATHENTRY- Flag indicating bathroom on the entry level (when there's more than 1 floor)
HABEDENTRY - Flag indicating bedroom on the entry level (when there's more than 1 floor)
HALIFT- Flag indicating home has a chairlift, stairlift, or platform lift
HARAMP- Flag indicating home has a ramp
BLD - Type of housing unit
<!-- Response Codes (BLD) -->
<!-- 01: Mobile home or trailer -->
<!-- 02: One-family house, detached -->
<!-- 03: One-family house, attached -->
<!-- 04: 2 apartments -->
<!-- 05: 3 to 4 apartments -->
<!-- 06: 5 to 9 apartments -->
<!-- 07: 10 to 19 apartments -->
<!-- 08: 20 to 49 apartments -->
<!-- 09: 50 or more apartments -->
<!-- 10: Boat, RV, van, etc. -->
YRBUILT <- Year unit was built
<!-- Response Codes -->
<!-- 2010 : YYYY: 2010 to YYYY -->
<!-- 2000: 2000 to 2009 -->
<!-- 1990: 1990 to 1999 -->
<!-- 1980: 1980 to 1989 -->
<!-- 1970: 1970 to 1979 -->
<!-- 1960: 1960 to 1969 -->
<!-- 1950: 1950 to 1959 -->
<!-- 1940: 1940 to 1949 -->
<!-- 1930: 1930 to 1939 -->
<!-- 1920: 1920 to 1929 -->
<!-- 1919: 1919 or earlier -->
MARKETVAL <- Current market value of unit

```{r functions}

#initialize funcs used throughout 
w_pct <- function(x, w) {
  weighted.mean(x, w, na.rm = TRUE) * 100
}

summarize_accessibility <- function(df) {
  df %>%
    summarize(
      no_step_pct       = w_pct(no_step == 1, WEIGHT),
      entry_bath_pct    = w_pct(entry_bath == 1, WEIGHT),
      entry_bed_pct     = w_pct(entry_bed == 1, WEIGHT),
      lift_pct          = w_pct(lift == 1, WEIGHT),
      ramp_pct          = w_pct(ramp == 1, WEIGHT),
      
      bed_bath_entry_pct = w_pct(entry_bath == 1 & entry_bed == 1, WEIGHT),
      
      core_accessibility_pct = w_pct(
        no_step == 1 & entry_bath == 1 & entry_bed == 1 |
        entry_bed == 1 & entry_bath == 1 & lift == 1,
        WEIGHT
      )
    )
}

```

```{r load_AHS}
library(glptools)
glp_load_packages(T)

AHS_2019_micro <- read.csv("C:/Users/morga/OneDrive - The Community Foundation of Louisville, Inc/data/ahs2019n.csv")
```

```{r clean_filter_AHS}

# 1: Yes
# 2: No
# M or -9: Not reported
# N or -6: Not applicable

accessible_features <- AHS_2019_micro %>%
  select(NOSTEP, UNITFLOORS, HABATHENTRY, HABEDENTRY, HALIFT, HARAMP, WEIGHT, BLD, YRBUILT, MARKETVAL, HASUPP, WCHAIR, ECHAIR, CANE, CRUTCH, MOBOTHER) %>%
  mutate(across(everything(), ~ as.integer(str_replace_all(str_trim(.), "'", "")))) %>%
  mutate(
    no_step = case_when(
      NOSTEP == 9 ~ NA, 
      NOSTEP == 1 ~ T,
      NOSTEP == 2 ~ F),
    
    entry_bath = case_when(
      HABATHENTRY == 9 ~ NA, 
      HABATHENTRY == 1 ~ T,
      HABATHENTRY == 2 ~ F, 
      HABATHENTRY == 6 ~ T # single floor unit 
      ), 
    
    entry_bed = case_when(
      HABEDENTRY == 9 ~ NA, 
      HABEDENTRY == 1 ~ T,
      HABEDENTRY == 2 ~ F, 
      HABEDENTRY == 6 ~ T # single floor unit 
      ), 
    
    lift = case_when(
      HALIFT == 9 ~ NA, 
      HALIFT == 1 ~ T,
      HALIFT == 2 ~ F, 
      HALIFT == 6 ~ T # single floor unit or not applicable  
      ), 
    
    ramp = case_when(
      HARAMP == 9 ~ NA, 
      HARAMP == 1 ~ T,
      HARAMP == 2 ~ F, 
      HARAMP == 6 ~ T # single floor unit or not applicable  
      ), 
    
    floors = case_when(
      UNITFLOORS == 1 ~ "1",
      UNITFLOORS == 2 ~ "2",
      UNITFLOORS == 3 ~ "3 or more"
    ),
    
    SF = BLD %in% c(1, 2, 3), 
    MF = BLD %in% 4:9,
    
    yr_built = case_when(
      YRBUILT <= 1940 ~ "built_1939_or_earlier" , 
      YRBUILT %in% 1940:1950 ~ "built_1940_to_1959", 
      YRBUILT %in% 1960:1970 ~ "built_1960_to_1979", 
      YRBUILT %in% 1980:1990 ~ "built_1980_to_1999", 
      YRBUILT <= 2000 ~ "built after 2000"
    ), 
    
    market_val = case_when( 
      MARKETVAL >= 9000000  ~ "more_than_10M", 
      MARKETVAL %in% 5000000:9000000 ~ "between_5_and_9M", 
      MARKETVAL %in% 1000000:4999999 ~ "between_1_and_5M", 
      MARKETVAL %in% 750000:999999 ~ "between_750k_and_1M", 
      MARKETVAL %in% 500000:749999 ~ "between_500k_and_750k",
      MARKETVAL %in% 300000:499999 ~ "between_300k_and_500k",
      MARKETVAL %in% 200000:299999 ~ "between_200k_and_300k",
      MARKETVAL %in% 100000:199999 ~ "between_100k_and_200k",
      MARKETVAL <= 99999 ~ "less than 100k"
      ), 
    
    HASUPP = case_when(
      HASUPP == -6 ~ NA, 
      HASUPP == -9 ~ NA, 
      HASUPP %in% c(1,2,3,4,5) ~ HASUPP
    )
)
save(accessible_features, file = "interim_data/accessible_features_hh.RData")
#rm(AHS_2019_micro)
```

```{r accessibility_summary}
library(readxl)
library(dplyr)
library(tidyr)

accessibility_summary_SF <- accessible_features %>%
  filter(SF) %>%
  #all individual features
  summarize_accessibility()

accessibility_summary_SF <- accessibility_summary_SF %>%
  pivot_longer(everything(), names_to = "feature", values_to = "percent")
  
save(accessibility_summary_SF, file = "interim_data/accessibility_summary_SF.RData")

jeff_co_SF <- read_excel("data/ACS_jeff_co_SF_estimates.xlsx") %>%
  filter(type == 'SF')

jeff_co_acc_SF_projections <- jeff_co_SF %>%
  crossing(accessibility_summary_SF) %>%     #every year × every metric
  mutate(
    #round down to nearest whole num
    estimated_SF_units = floor(units * (percent / 100))
  )%>%
  select(4:6)

save(jeff_co_acc_SF_projections, file = "interim_data/accessible_SF_housing_estimates.RData")
rm(jeff_co_SF, accessibility_summary_SF)
```

```{r}
accessibility_summary_MF <- accessible_features %>%
  filter(MF) %>%
  #all individual features
  summarize_accessibility()

accessibility_summary_MF <- accessibility_summary_MF %>%
  pivot_longer(everything(), names_to = "feature", values_to = "percent")

jeff_co_MF <- read_excel("data/ACS_jeff_co_SF_estimates.xlsx") %>%
  filter(type == 'MF')

jeff_co_acc_MF_projections <- jeff_co_MF %>%
  crossing(accessibility_summary_MF) %>%     #every year × every metric
  mutate(
    #round down to nearest whole num
    estimated_MF_units = floor(units * (percent / 100))
  )%>%
  select(4:6)

save(jeff_co_acc_MF_projections, file = "interim_data/accessible_MF_housing_estimates.RData")
rm(jeff_co_MF, accessibility_summary_MF)
```

```{r summary_by_year_built}
accessibility_by_yr_built <- accessible_features %>%
  select(no_step:ramp, yr_built, WEIGHT, SF) %>%
  filter(!is.na(yr_built), SF)%>%
  group_by(yr_built) %>%
  summarize_accessibility()%>%
  ungroup() %>%
  #order
  mutate(yr_built = yr_built <- factor(c("built after 2000", "built_1980_to_1999", "built_1960_to_1979", "built_1940_to_1959", "built_1939_or_earlier")))

accessibility_by_yr_built<- mutate(accessibility_by_yr_built, yr_built = as.character(yr_built))

jeff_co_estimates_by_yr <- read_excel("data/jeff_co_estimates_by_yr.xlsx") 

jeff_co_estimates_by_yr <- jeff_co_estimates_by_yr%>%
  rename(yr_built = built)

accessibility_by_yr_built_projections <- jeff_co_estimates_by_yr %>%
  left_join(accessibility_by_yr_built, by = "yr_built") %>%
  pivot_longer(no_step_pct:core_accessibility_pct, names_to = "acc_measure", values_to = "percent") %>%
  mutate(
    estimated_units = floor(units * percent / 100)
)
rm(jeff_co_estimates_by_yr, accessibility_by_yr_built)
```

```{r summary_by_property_value}

# no available property val var in ACS to project estimated units with 
accessibility_by_property_val <- accessible_features %>%
  select(no_step:ramp, market_val, WEIGHT, SF) %>%
  filter(!is.na(market_val), SF) %>%
  group_by(market_val) %>%
  summarize_accessibility() %>%
  ungroup()%>%
  mutate(market_val = market_val <- factor(c("more_than_10M", "between_5_and_9M", "between_1_and_5M", "between_750k_and_1M", "between_500k_and_750k", "between_300k_and_500k", "between_200k_and_300k", "between_100k_and_200k", "less than 100k")))
```

```{r mobility_device}

#create flag for person who uses mobility device
accessible_features %<>%
  mutate(across(c(WCHAIR, ECHAIR, CANE, CRUTCH, MOBOTHER),
    ~ case_when(.x %in% c(-6, -9) ~ NA, TRUE ~ .x)
  )) %>%
  mutate(any_mob_dev = WCHAIR == 1 |ECHAIR == 1 | CANE == 1 | CRUTCH == 1 | MOBOTHER == 1)

#calculate % ppl who use mobility device 
mob_dev_pct <- accessible_features %>%
  filter(!is.na(any_mob_dev)) %>%
  summarize(percent_mob_dev = w_pct(any_mob_dev, WEIGHT))

#Electric Wheelchair vs Manual
wchair_pct <- accessible_features %>%
  filter(!is.na(WCHAIR)) %>%
  summarise(wchair_pct = w_pct(WCHAIR == 1, WEIGHT)) # = 1.60573 %

echair_pct <- accessible_features %>%
  filter(!is.na(ECHAIR)) %>%
  summarise(echair_pct = w_pct(ECHAIR == 1, WEIGHT)) # = 	4.121627 %
```

```{r resident_ratings}

# compare respondent rating (housing fit) with those units w "core accessibility" 
# core accessibility = no step, bed, bath OR lift, bed, bath

#create core acc flag to sort by later
accessible_features <- accessible_features %>%
  mutate(
    core_accessibility = (no_step == 1 & entry_bath == 1 & entry_bed == 1) |
                         (entry_bed == 1 & entry_bath == 1 & lift == 1), 
    no_accessibility = (no_step == 0 & entry_bath == 0 & entry_bed == 0 & lift == 0 & ramp == 0)
  )

# SF
# respondent rating for units which have "core accessibility" 
RR_summary <- accessible_features %>%
  filter(SF, core_accessibility, !is.na(HASUPP)) %>%
  summarise(
    rating1_pct = w_pct(HASUPP == "1", WEIGHT),
    rating2_pct = w_pct(HASUPP == "2", WEIGHT),
    rating3_pct = w_pct(HASUPP == "3", WEIGHT),
    rating4_pct = w_pct(HASUPP == "4", WEIGHT),
    rating5_pct = w_pct(HASUPP == "5", WEIGHT)
  ) %>%
  pivot_longer(everything(), names_to = "rating", values_to = "core_percent" ) %>%
  mutate(rating = row_number())

# respondent rating for units which do not have "core accessibility"
not_acc_RR_summary <- accessible_features %>%
  filter(SF, !core_accessibility, !is.na(HASUPP)) %>%
  summarise(
    rating1_pct = w_pct(HASUPP == "1", WEIGHT),
    rating2_pct = w_pct(HASUPP == "2", WEIGHT),
    rating3_pct = w_pct(HASUPP == "3", WEIGHT),
    rating4_pct = w_pct(HASUPP == "4", WEIGHT),
    rating5_pct = w_pct(HASUPP == "5", WEIGHT)
  ) %>%
  pivot_longer(everything(), names_to = "rating", values_to = "not_core_percent" ) %>%
  mutate(rating = row_number())

left_join(RR_summary, not_acc_RR_summary, by = "rating")

# same categories but for MF
MF_RR <- accessible_features %>%
  filter(MF, core_accessibility, !is.na(HASUPP)) %>%
  summarise(
    rating1_pct = w_pct(HASUPP == "1", WEIGHT),
    rating2_pct = w_pct(HASUPP == "2", WEIGHT),
    rating3_pct = w_pct(HASUPP == "3", WEIGHT),
    rating4_pct = w_pct(HASUPP == "4", WEIGHT),
    rating5_pct = w_pct(HASUPP == "5", WEIGHT)
  ) %>%
  pivot_longer(everything(), names_to = "rating", values_to = "MF_core_percent" ) %>%
  mutate(rating = row_number())

MF_RR_not_core <- accessible_features %>%
  filter(MF, !core_accessibility, !is.na(HASUPP)) %>%
  summarise(
    rating1_pct = w_pct(HASUPP == "1", WEIGHT),
    rating2_pct = w_pct(HASUPP == "2", WEIGHT),
    rating3_pct = w_pct(HASUPP == "3", WEIGHT),
    rating4_pct = w_pct(HASUPP == "4", WEIGHT),
    rating5_pct = w_pct(HASUPP == "5", WEIGHT)
  ) %>%
  pivot_longer(everything(), names_to = "rating", values_to = "MF_not_core_percent" ) %>%
  mutate(rating = row_number())

left_join(MF_RR, MF_RR_not_core, by = "rating")

# NO accessibility features 
strict_RR_summary <- accessible_features %>%
  filter(SF, no_accessibility, !is.na(HASUPP), !is.na(any_mob_dev)) %>%
  summarise(
    rating1_pct = w_pct(HASUPP == "1", WEIGHT),
    rating2_pct = w_pct(HASUPP == "2", WEIGHT),
    rating3_pct = w_pct(HASUPP == "3", WEIGHT),
    rating4_pct = w_pct(HASUPP == "4", WEIGHT),
    rating5_pct = w_pct(HASUPP == "5", WEIGHT)
  ) %>%
  pivot_longer(everything(), names_to = "rating", values_to = "core_percent" ) %>%
  mutate(rating = row_number())
```


```{r}
# looking at common traits for those who gave low satisfaction scores 
dig_dig_dig <- accessible_features %>%
  filter(HASUPP == 1 | HASUPP == 2) 
  # summarise(
  #   pct_no_step     = weighted.mean(no_step, WEIGHT, na.rm = TRUE),
  #   pct_entry_bath  = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE),
  #   pct_entry_bed   = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE),
  #   pct_lift        = weighted.mean(lift, WEIGHT, na.rm = TRUE),
  #   pct_ramp        = weighted.mean(ramp, WEIGHT, na.rm = TRUE),
  #   most_common_floors      = floors[which.max(table(floors))],
  #   most_common_market_val  = market_val[which.max(table(market_val))],
  #   most_common_yr_built    = yr_built[which.max(table(yr_built))]
  # )

```



