---
title: "projections"
output: html_document
---
###2019 AHS Variables Measuring Accessibility: 
NOSTEP - Flag indicating unit's entrance is accessible without climbing up or down any stairs
UNITFLOORS - Number of floors within the unit
HABATHENTRY- Flag indicating bathroom on the entry level (when there's more than 1 floor)
HABEDENTRY - Flag indicating bedroom on the entry level (when there's more than 1 floor)
HALIFT- Flag indicating home has a chairlift, stairlift, or platform lift
HARAMP- Flag indicating home has a ramp
BLD - Type of housing unit
<!-- Response Codes (BLD) -->
<!-- 01: Mobile home or trailer -->
<!-- 02: One-family house, detached -->
<!-- 03: One-family house, attached -->
<!-- 04: 2 apartments -->
<!-- 05: 3 to 4 apartments -->
<!-- 06: 5 to 9 apartments -->
<!-- 07: 10 to 19 apartments -->
<!-- 08: 20 to 49 apartments -->
<!-- 09: 50 or more apartments -->
<!-- 10: Boat, RV, van, etc. -->
YRBUILT <- Year unit was built
<!-- Response Codes -->
<!-- 2010 : YYYY: 2010 to YYYY -->
<!-- 2000: 2000 to 2009 -->
<!-- 1990: 1990 to 1999 -->
<!-- 1980: 1980 to 1989 -->
<!-- 1970: 1970 to 1979 -->
<!-- 1960: 1960 to 1969 -->
<!-- 1950: 1950 to 1959 -->
<!-- 1940: 1940 to 1949 -->
<!-- 1930: 1930 to 1939 -->
<!-- 1920: 1920 to 1929 -->
<!-- 1919: 1919 or earlier -->
MARKETVAL <- Current market value of unit


```{r load_AHS}

library(glptools)
glp_load_packages(T)

AHS_2019_micro <- read.csv("C:/Users/morga/OneDrive - The Community Foundation of Louisville, Inc/data/ahs2019n.csv")
```

```{r clean_filter_AHS}

# 1: Yes
# 2: No
# M or -9: Not reported
# N or -6: Not applicable

accessible_features <- AHS_2019_micro %>%
  select(NOSTEP, UNITFLOORS, HABATHENTRY, HABEDENTRY, HALIFT, HARAMP, WEIGHT, BLD, YRBUILT, MARKETVAL) %>%
  mutate(across(everything(), ~ as.integer(str_replace_all(str_trim(.), "'", "")))) %>%
  mutate(
    no_step = case_when(
      NOSTEP == 9 ~ NA, 
      NOSTEP == 1 ~ T,
      NOSTEP == 2 ~ F),
    
    entry_bath = case_when(
      HABATHENTRY == 9 ~ NA, 
      HABATHENTRY == 1 ~ T,
      HABATHENTRY == 2 ~ F, 
      HABATHENTRY == 6 ~ T # single floor unit 
      ), 
    
    entry_bed = case_when(
      HABEDENTRY == 9 ~ NA, 
      HABEDENTRY == 1 ~ T,
      HABEDENTRY == 2 ~ F, 
      HABEDENTRY == 6 ~ T # single floor unit 
      ), 
    
    lift = case_when(
      HALIFT == 9 ~ NA, 
      HALIFT == 1 ~ T,
      HALIFT == 2 ~ F, 
      HALIFT == 6 ~ T # single floor unit or not applicable  
      ), 
    
    ramp = case_when(
      HARAMP == 9 ~ NA, 
      HARAMP == 1 ~ T,
      HARAMP == 2 ~ F, 
      HARAMP == 6 ~ T # single floor unit or not applicable  
      ), 
    
    floors = case_when(
      UNITFLOORS == 1 ~ "1",
      UNITFLOORS == 2 ~ "2",
      UNITFLOORS == 3 ~ "3 or more"
    ),
    
    SF = BLD %in% c(1, 2, 3), 
    MF = BLD %in% 4:9,
    
    yr_built = case_when(
      YRBUILT <= 1940 ~ "built_1939_or_earlier" , 
      YRBUILT %in% 1940:1950 ~ "built_1940_to_1959", 
      YRBUILT %in% 1960:1970 ~ "built_1960_to_1979", 
      YRBUILT %in% 1980:1990 ~ "built_1980_to_1999", 
      YRBUILT <= 2000 ~ "built after 2000"
    ), 
    
    market_val = case_when( 
      MARKETVAL >= 9000000  ~ "more_than_10M", 
      MARKETVAL %in% 5000000:9000000 ~ "between_5_and_9M", 
      MARKETVAL %in% 1000000:4999999 ~ "between_1_and_5M", 
      MARKETVAL %in% 750000:999999 ~ "between_750k_and_1M", 
      MARKETVAL %in% 500000:749999 ~ "between_500k_and_750k",
      MARKETVAL %in% 300000:499999 ~ "between_300k_and_500k",
      MARKETVAL %in% 200000:299999 ~ "between_200k_and_300k",
      MARKETVAL %in% 100000:199999 ~ "between_100k_and_200k",
      MARKETVAL <= 99999 ~ "less than 100k"
      )
)
save(accessible_features, file = "interim_data/accessible_features_hh.RData")
```

```{r accessibility_summary}
accessibility_summary_SF <- accessible_features %>%
  filter(SF) %>%
  #all individual features
  summarize(
    no_step_pct = weighted.mean(no_step, WEIGHT, na.rm = TRUE) * 100,
    entry_bath_pct = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE) * 100,
    entry_bed_pct = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE) * 100,
    lift_pct = weighted.mean(lift, WEIGHT, na.rm = TRUE) * 100,
    ramp_pct = weighted.mean(ramp, WEIGHT, na.rm = TRUE) * 100, 
  
    bed_bath_entry = weighted.mean(
      entry_bath == 1 & entry_bed == 1,
      WEIGHT, 
      na.rm = TRUE
    ) * 100, 
    
    #basic accessibility
    core_accessibility_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1,
    WEIGHT,
    na.rm = TRUE
  ) * 100,

  # basic features plus ramp OR lift
  core_plus_ramp_lift_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1 & (ramp == 1 | lift == 1),
    WEIGHT,
    na.rm = TRUE
  ) * 100
  )

accessibility_summary_SF <- accessibility_summary_SF %>%
  pivot_longer(everything(), names_to = "feature", values_to = "percent")
  
save(accessibility_summary_SF, file = "interim_data/accessibility_summary_SF.RData")
```

```{r}
accessibility_summary_MF <- accessible_features %>%
  filter(MF) %>%
  #all individual features
  summarize(
    no_step_pct = weighted.mean(no_step, WEIGHT, na.rm = TRUE) * 100,
    entry_bath_pct = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE) * 100,
    entry_bed_pct = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE) * 100,
    lift_pct = weighted.mean(lift, WEIGHT, na.rm = TRUE) * 100,
    ramp_pct = weighted.mean(ramp, WEIGHT, na.rm = TRUE) * 100, 
  
    bed_bath_entry = weighted.mean(
      entry_bath == 1 & entry_bed == 1,
      WEIGHT, 
      na.rm = TRUE
    ) * 100, 
    
    #basic accessibility
    core_accessibility_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1,
    WEIGHT,
    na.rm = TRUE
  ) * 100,

  # basic features plus ramp OR lift
  core_plus_ramp_lift_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1 & (ramp == 1 | lift == 1),
    WEIGHT,
    na.rm = TRUE
  ) * 100
  )

accessibility_summary_MF <- accessibility_summary_MF %>%
  pivot_longer(everything(), names_to = "feature", values_to = "percent")
  
save(accessibility_summary_MF, file = "interim_data/accessibility_summary_MF.RData")

```

```{r summary_by_year_built}

accessibility_by_yr_built <- accessible_features %>%
  select(no_step:ramp, yr_built, WEIGHT) %>%
  filter(!is.na(yr_built))%>%
  group_by(yr_built) %>%
  summarise(
    pct_no_step = weighted.mean(no_step, WEIGHT, na.rm = TRUE),
    pct_entry_bath = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE),
    pct_entry_bed = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE),
    pct_lift = weighted.mean(lift, WEIGHT, na.rm = TRUE),
    pct_ramp = weighted.mean(ramp, WEIGHT, na.rm = TRUE), 
    
    bed_bath_entry = weighted.mean(
      entry_bath == 1 & entry_bed == 1,
      WEIGHT, 
      na.rm = TRUE
    ) * 100, 
    
    #basic accessibility
    core_accessibility_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1,
    WEIGHT,
    na.rm = TRUE
  ) * 100,

  # basic features plus ramp OR lift
  core_plus_ramp_lift_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1 & (ramp == 1 | lift == 1),
    WEIGHT,
    na.rm = TRUE
  ) * 100
  ) %>%
  ungroup() %>%
  #order
  mutate(yr_built = yr_built <- factor(c("built after 2000", "built_1980_to_1999", "built_1960_to_1979", "built_1940_to_1959", "built_1939_or_earlier")))
```

```{r summary_by_property_value}

accessibility_by_property_val <- accessible_features %>%
  select(no_step:ramp, market_val, WEIGHT) %>%
  filter(!is.na(market_val)) %>%
  group_by(market_val) %>%
  summarise(pct_no_step = weighted.mean(no_step, WEIGHT, na.rm = TRUE),
    pct_entry_bath = weighted.mean(entry_bath, WEIGHT, na.rm = TRUE),
    pct_entry_bed = weighted.mean(entry_bed, WEIGHT, na.rm = TRUE),
    pct_lift = weighted.mean(lift, WEIGHT, na.rm = TRUE),
    pct_ramp = weighted.mean(ramp, WEIGHT, na.rm = TRUE), 
    
    bed_bath_entry = weighted.mean(
      entry_bath == 1 & entry_bed == 1,
      WEIGHT, 
      na.rm = TRUE
    ) * 100, 
    
    #basic accessibility
    core_accessibility_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1,
    WEIGHT,
    na.rm = TRUE
  ) * 100,

  # basic features plus ramp OR lift
  core_plus_ramp_lift_pct = weighted.mean(
    no_step == 1 & entry_bath == 1 & entry_bed == 1 & (ramp == 1 | lift == 1),
    WEIGHT,
    na.rm = TRUE
  ) * 100
  ) %>%
  ungroup()%>%
  mutate(market_val = market_val <- factor(c("more_than_10M", "between_5_and_9M", "between_1_and_5M", "between_750k_and_1M", "between_500k_and_750k", "between_300k_and_500k", "between_200k_and_300k", "between_100k_and_200k", "less than 100k")))
  
  

```

```{r ACS_housing_estimates}
library(readxl)
library(dplyr)
library(tidyr)

jeff_co_HU <- read_excel("data/CO-EST2024-HU-21.xlsx")

jeff_co_HU <- jeff_co_HU %>%
  janitor::row_to_names(4, remove_rows_above = TRUE) %>%
  janitor::clean_names() %>%
  rename(estimates_base = april_1_2020_estimates_base) %>%
  mutate(estimates_base = as.integer(estimates_base)) %>%
  filter(geographic_area == ".Jefferson County, Kentucky") %>%
  pivot_longer(estimates_base:x2024, names_to = "year", values_to = "total_units")

```

```{r project!}
jeff_acc_housing_est <- jeff_co_HU %>%
  filter(year == "x2024") %>% #most recent estimates
  crossing(accessibility_summary) %>%     #every year × every metric
  mutate(
    #round down to nearest whole num
    estimated_units = floor(total_units * (percent / 100))
  ) 

save(jeff_acc_housing_est, file = "interim_data/accessible_housing_estimates.RData")
```
